Density-dependent selection and the limits of relative fitness

### 2018-04-26 manuscript.tex
l

Selection is commonly described by assigning relative fitness values to genotypes. Yet many populations are density-regulated, and the strength of selection can then depend on density. Additionally, selection can induce changes in density. When these two effects occur together, the resulting selection dynamics are not describable by density-independent relative fitness values, particularly if selection is strong. In most previous models of density-dependent selection, these effects are assumed to occur together (e.g. ``$K$-selection’’), and as a result relative fitness models seem to require weak selection. Here we argue that constant relative fitness models are valid under a substantially broader range of conditions. We develop a novel model of density-regulated population growth with three traits: fecundity, mortality, and competitive ability. Our model contains a finite ``reproductive excess’’, which allows selection on competitive ability to occur without affecting density. Remarkably, we find that even strong selection is compatible with density-independent relative fitnesses provided that only one trait evolves at a time. Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models. In such cases our model offers a possible alternative to relative fitness.

\noindent (187 words)

>>>>

Selection is commonly described by assigning constant relative fitness values to genotypes. Yet population density is often regulated by crowding. Relative fitness may then depend on density, and density will change when selection acts on a density-regulating trait. When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically-stable populations. In most previous models of density-dependent selection, density-regulating traits are subject to density-dependent selection (e.g. ``$K$-selection’’), suggesting that density-independent relative fitnesses must be replaced with more ecologically-explicit absolute fitnesses unless selection is weak. Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection. We develop a novel model of density-regulated population growth with three traits: fecundity, mortality, and competitive ability. Our model incorporates a density-dependent juvenile ``reproductive excess’’, which largely decouples density-dependent selection from the regulation of density. We find that density-independent relative fitnesses accurately describe strong selection on each trait separately, even for fecundity, which is both density-regulating and subject to density-dependent selection. Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models. In such cases our model offers a possible alternative to relative fitness.

\noindent (197 words)
